### 2013-07-23 doc/asplos14/memory.tex
many small fixes

Applications written for \Grappa utilize two forms of memory: local and
global. Local memory is local to a single core within a node in the system.
Accesses occur through conventional pointers. Applications use local accesses for a
number of things in \Grappa: the stack associated with a task, accesses to
localized global memory in caches (see below), and accesses to debugging
infrastructure that is local to each system node. Local pointers cannot access
memory on other cores, and are valid only on their home core.

@@

\paragraph{Global memory addressing} \Grappa provides two methods for storing
data in the global memory. The first is a distributed heap striped across all
the machines in the system in a block cyclic fashion. The
\texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
and deallocate memory in the global heap. Addresses to memory in the global
heap use \emph{linear addresses}. Choosing the block size involves trading off
sequential bandwidth against aggregate random access bandwidth. Smaller block
sizes help spread data across all the memory controllers in the cluster, but
larger block sizes allow the locality-optimized memory controllers to provide
increased sequential bandwidth. The block size, which is configurable, is
typically set to 64 bytes, or the size of a single hardware cache line, in
order to exploit spatial locality when available.

\Grappa also allows any local data on a core's stacks or heap to be exported
to the global address space to be made accessible to other cores across the
system. Addresses to global memory allocated in this way use \emph{2D global
addresses}. This uses a traditional PGAS (partitioned global address
space~\cite{upc:2005}) addressing model, where each address is a tuple of a
rank in the job (or global process ID) and an address in that process. The
lower 48 bits of the address hold a virtual address in the process. The top
bit is set to indicate that the reference is a 2D address (as opposed to
linear address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made accessible
to other cores in the system by wrapping the address and node ID into a 2D
global address. This address can then be accessed with a delegate operation
and even be buffered by other cores. The address is converted at the destination
into a canonical x86 address by replacing the upper bits with the
sign-extended upper bit of the virtual address. 2D addresses may refer to
memory allocated from a single processes' heap or from a task's stack.
Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can refer
to other cores' memory.

@@

\paragraph{Global memory access} 
Access to \Grappa's distributed shared memory is provided through  {\em
delegate} operations, which are short memory accesses performed at the memory
location's home node. When the data access pattern has
low-locality, it is more efficient to modify the data on its home core rather
than bringing a copy to the requesting core and returning it after
modification. Delegate operations~\cite{Nelson:hotpar11, delegated:oopsla11}
provide this capability. Applications can dispatch computation to be performed
on individual machine-word sized chunks of global memory to the memory system
itself. Delegates can execute arbitrary non-blocking code, so we use them to perform simple
\emph{read\/}/\emph{write\/} operations to global memory, as well as more complex \emph{read-modify-write\/} operations (e.g., \emph{fetch-and-add\/}).

@@

Delegate operations can be either {\em blocking} or {\em non-blocking}.
With blocking operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in \Grappa
is implemented using blocking delegate operations. On the other hand, remote data
accesses often can overlap. In order to avoid
unnecessary waiting, we support non-blocking delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by non-blocking delegate
invocations. Delegate write operations may also be performed as
non-blocking,
and since they do not return data, a mechanism to detect completion is needed.
\Grappa provides a \texttt{GlobalCompletionEvent} synchronization object, which non-blocking operations (including tasks) can be enrolled with and which other tasks can block on to be notified and woken when all enrolled operations in the collection are complete.
\TODO{augment this with our bulk futures ie global completions}

>>>>

Applications written for \Grappa utilize two forms of memory: local and
global. Local memory is local to a single core within a node in the system.
Accesses occur through conventional pointers. Applications use local accesses for a
number of things in \Grappa: the stack associated with a task, accesses to
localized global memory in caches (see below), and accesses to debugging
infrastructure local to each system node. Local pointers cannot access
memory on other cores, and are valid only on their home core.

@@

\paragraph{Global memory addressing} \Grappa provides two methods for storing
data in the global memory. The first is a distributed heap striped across all
the machines in the system in a block-cyclic fashion. The
\texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
and deallocate memory in the global heap. Addresses to memory in the global
heap use \emph{linear addresses}. Choosing the block size involves trading off
sequential bandwidth against aggregate random access bandwidth. Smaller block
sizes help spread data across all the memory controllers in the cluster, but
larger block sizes allow the locality-optimized memory controllers to provide
increased sequential bandwidth. The block size, which is configurable, is
typically set to 64 bytes, or the size of a single hardware cache line, in
order to exploit spatial locality when available.

\Grappa also allows any local data on a core's stacks or heap to be exported
to the global address space to be made accessible to other cores across the
system. Addresses to global memory allocated in this way use \emph{2D global
addresses}. This uses a traditional PGAS (partitioned global address
space~\cite{upc:2005}) addressing model, where each address is a tuple of a
rank in the job (or global process ID) and an address in that process. The
lower 48 bits of the address hold a virtual address in the process. The top
bit is set to indicate that the reference is a 2D address (as opposed to
linear address). This leaves 15~bits for network endpoint ID, which limits our
scalability to $2^{15}$ cores. Any node-local data can be made accessible
to other cores in the system by wrapping the address and node ID into a 2D
global address. This address can then be accessed with a delegate operation
and even be buffered by other cores. The address is converted at the destination
into a canonical x86 address by replacing the upper bits with the
sign-extended upper bit of the virtual address. 2D addresses may refer to
memory allocated from a single processes' heap or from a task's stack.
Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can refer
to other cores' memory.

@@

\paragraph{Global memory access} 
Access to \Grappa's distributed shared memory is provided through  {\em
delegate\/} operations, which are short memory accesses performed at the memory
location's home node. When the data access pattern has
low-locality, it is more efficient to modify the data on its home core rather
than bringing a copy to the requesting core and returning it after
modification. Delegate operations~\cite{Nelson:hotpar11, delegated:oopsla11}
provide this capability. Applications can dispatch computation to be performed
on individual machine-word sized chunks of global memory to the memory system
itself. Delegates can execute arbitrary non-blocking code, so we use them to perform simple
\emph{read\/}/\emph{write\/} operations to global memory, as well as more complex \emph{read-modify-write\/} operations (e.g., \emph{fetch-and-add\/}).

@@

Delegate operations can be either {\em blocking\/} or {\em non-blocking}.
With blocking operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in \Grappa
is implemented using blocking delegate operations. On the other hand, remote data
accesses often can overlap. In order to avoid
unnecessary waiting, we support non-blocking delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by non-blocking delegate
invocations. Delegate write operations may also be performed as
non-blocking,
and since they do not return data, a mechanism to detect completion is needed.
\Grappa provides a \texttt{GlobalCompletionEvent} synchronization object, which non-blocking operations (including tasks) can be enrolled with and which other tasks can block on to be notified and woken when all enrolled operations in the collection are complete.
\TODO{augment this with our bulk futures ie global completions}

### 2013-07-23 doc/asplos14/method.tex
many small fixes

We implemented the \Grappa in C++ for the Linux operating system. The core
runtime system system is about 15K lines of code. We ported a number of
benchmarks to \Grappa as well as collected and optimized a set of comparison
benchmarks for XMT, MPI and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand
network cards. Nodes are connected via a QLogic InfiniBand switch. We also
compare \Grappa to a 128-node Cray XMT (3rd generation MTA). Each node
consists of a 500-MHz MTA Threadstorm multithreaded processor that supports
128~streams. The machine uses Cray's proprietary SeaStar2 interconnection
network.

@@

\vspace{0.5ex}
\noindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
Graph500 benchmark and is what currently determines the ranking of machines on
the Graph500 list~\cite{graph500list}. As a whole, the Graph500 benchmark
suite is designed to bring the focus of system design on data-intensive
workloads, particularly large-scale graph analysis problems, that are
important among cybersecurity, informatics, and network understanding
workloads. The BFS benchmark builds a search tree containing parent nodes for
each traversed vertex during the search. While this is a relatively simple
problem to solve, it exercises the random-access and fine-grained
synchronization capabilities of a system as well as being a primitive in many
other graph algorithms. Performance is measured in \emph{traversed edges per
second\/} (TEPS), where the number of edges is the edges making up the
generated BFS tree. With some modifications to the XMT reference version of Graph500 BFS,
the XMT compiler can be made to recognize and apply a Manhattan loop collapse, exposing enough parallelism to allow it to scale out to 64 nodes for the problem scales we show.
We do not employ algorithmic improvements for any of these versions, though there are
many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}. \Grappa can be expected to benefit the same as MPI due to decreased communication.

\vspace{0.5ex}
\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the CrayXMT's early predecessor once held the world speed record~\cite{TeraRecord}. The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system. Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets. For these reasons, the MPI reference version and our \Grappa implementation use bucket sort. On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.  The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS). For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS''.

\vspace{0.5ex}
\noindent{\bf PageRank:} This is a common centrality metric for
graphs. Computing PageRank is an iterative algorithm with a common
pattern of gather, apply, and scatter on the rank of vertex. The algorithm
is often implemented by sparse linear algebra libraries, with the main
kernel being the sparse matrix dense vector multiply. For the multiply
step, \Grappa parallelizes over the rows and parallelizes each dot
product. PageRank has the fortunate property that the accumulation
function over the in-edges is associative and commutative, so they can
be processed in any order in parallel. Rather than the programmer writing the
parallel dot product as local accumulations with a final all-reduce
step, we simply send streaming increments to each element of the final
vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT \cite{MTGL:pagerank}.
For \Grappa, we run on a scale 29 graph using the Graph500 generator.

>>>>

We implemented the \Grappa in C++ for the Linux operating system. The core
runtime system system is about 15K lines of code. We ported a number of
benchmarks to \Grappa as well as collected and optimized a set of comparison
benchmarks for XMT, MPI, and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand
network cards. Nodes are connected via a QLogic InfiniBand switch. We also
compare \Grappa to a 128-node Cray XMT (3rd generation MTA). Each node
consists of a 500-MHz MTA Threadstorm multithreaded processor that supports
128~streams. The machine uses Cray's proprietary SeaStar2 interconnection
network.

@@

\vspace{0.5ex}
\noindent{\bf Breadth-first-search (BFS):} This is the primary kernel for the
Graph500 benchmark and is what currently determines the ranking of machines on
the Graph500 list~\cite{graph500list}. As a whole, the Graph500 benchmark
suite is designed to bring the focus of system design on data-intensive
workloads, particularly large-scale graph analysis problems, that are
important among cybersecurity, informatics, and network-understanding
workloads. The BFS benchmark builds a search tree containing parent nodes for
each traversed vertex during the search. While this is a relatively simple
problem to solve, it exercises the random-access and fine-grained
synchronization capabilities of a system as well as being a primitive in many
other graph algorithms. Performance is measured in \emph{traversed edges per
second\/} (TEPS), where the number of edges is the edges making up the
generated BFS tree. With some modifications to the XMT reference version of Graph500 BFS,
the XMT compiler can be made to recognize and apply a Manhattan loop collapse, exposing enough parallelism to allow it to scale out to 64 nodes for the problem scales we show.
We do not employ algorithmic improvements for any of these versions, though there are
many~\cite{Beamer:Graph500,Yoo:FixedPointGraph500}. \Grappa can be expected to benefit the same as MPI due to decreased communication.

\vspace{0.5ex}
\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the CrayXMT's early predecessor once held the world speed record~\cite{TeraRecord}. The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system. Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets. For these reasons, the MPI reference version and our \Grappa implementation use bucket sort. On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.  The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS). For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS.''

\vspace{0.5ex}
\noindent{\bf PageRank:} This is a common centrality metric for
graphs. Computing PageRank is an iterative algorithm with a common
pattern of gather, apply, and scatter on the rank of vertex. The algorithm
is often implemented by sparse linear algebra libraries, with the main
kernel being the sparse matrix dense vector multiply. For the multiply
step, \Grappa parallelizes over the rows and parallelizes each dot
product. PageRank has the fortunate property that the accumulation
function over the in-edges is associative and commutative, so they can
be processed in any order or in parallel. Rather than the programmer writing the
parallel dot product as local accumulations with a final all-reduce
step, we simply send streaming increments to each element of the final
vector. We compare PageRank to published results for the Trilinos linear algebra library implemented in MPI, and multithreaded PageRank for the XMT \cite{MTGL:pagerank}.
For \Grappa, we run on a scale 29 graph using the Graph500 generator.

### 2013-07-23 doc/asplos14/tasks.tex
many small fixes

The basic unit of execution in \Grappa is a {\em task}. When tasks are ready to
execute, they are mapped to a {\em worker}, which is akin to a user-level
thread. Each hardware core has a single operating system thread pinned to it.

@@

\item {\it private task queue}, a queue of tasks that must run on
this core and therefore is not subject to stealing;

@@

\end{itemize}

Whenever a task yields, the scheduler makes a decision about what
to do next. First, any task in the deadline task queue whose deadline
is imminent is chosen for execution. This queue manages high priority
system tasks, such as periodically servicing communication requests. Second,
the scheduler determines if any workers with running tasks are ready to
execute; if so, one is scheduled. Finally, if no workers are ready to
run, but tasks are waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.

@@

\Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second,
the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken:1992:AMM:139669.140382}; that is, to run a
small non-blocking piece of code on a particular core in the system without waiting for
 a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} call for this purpose.

>>>>

The basic unit of execution in \Grappa is a {\em task}. When tasks are ready to
execute, they are mapped to a {\em worker,} which is akin to a user-level
thread. Each hardware core has a single operating system thread pinned to it.

@@

\item {\it private task queue}, a queue of tasks that must run on
this core and is therefore not subject to stealing;

@@

\end{itemize}
Whenever a task yields, the scheduler makes a decision about what
to do next. First, any task in the deadline task queue whose deadline
is imminent is chosen for execution. This queue manages high priority
system tasks, such as periodically servicing communication requests. Second,
the scheduler determines if any workers with running tasks are ready to
execute; if so, one is scheduled. Finally, if no workers are ready to
run, but tasks are waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.

@@

\Grappa provides three methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second,
the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
small non-blocking piece of code on a particular core in the system without waiting for
 a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} routine for this purpose.
